A comprehensive analysis of natural sequence variation within the <i>Arabidopsis

### 2018-07-12 manuscript/04-IAAs.Rmd
-Cleaned up manuscript folder -moved to bookdown format -revised all sections

---
title: A comprehensive analysis of natural sequence variation within the <i>Arabidopsis
  thaliana</i> nuclear auxin signaling pathway using an accessible web application
author:
  affiliation: University of Washington
  name: R. Clay Wright
date: "`r Sys.Date()`"
params: 
  output: word_document
output:
  word_document:
    toc: yes
  html_document:
    fig_caption: yes
    toc: yes
always_allow_html: yes
fontfamily: Helvetica
fontsize: 11pt
keywords: r markdown
link-citations: yes
csl: apa.csl
bibliography:
- natural_variation.bib
- TIR1_AFB.bib
- ARF_C.bib
- AuxIAA_citations.bib
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, collapse = TRUE, comment = "#>", tidy = TRUE, cache = TRUE, cache.lazy = FALSE, out.width = "90%", fig.height=4, fig.width=7, out.extra='style="margin: auto; display: block; padding-top: 15px;"')

#devtools::install_github("wrightrc/r1001genomes", ref = "auxin-natural-variation")
library(ape)
library(r1001genomes)
library(DT)
library(tidyverse)
library(magrittr)
library(RColorBrewer)
library(ggpmisc)
library(ggthemes)
library(ggseqlogo)
library(ggtree)
library(reticulate)
library(scales)
library(viridis)
library(ips)
library(knitr)

#use_python(python = "/usr/local/bin/python")
```
<!--This is a CSS (Cascading Style Sheet).
 It controls how panels and images are formatted.></!-->
<style>
  img {
    display: block;
    margin: auto;
    width: 90%;
    height: auto;
  }
</style>

# List of working titles  
* A comprehensive analysis of natural sequence variation within the <i>Arabidopsis thaliana</i> nuclear auxin signaling pathway  
* **Add your suggestions**

# Abstract

* Motivating, high level statement  
* Statement defining the cutting edge of your research area  
* A statement of the form: "However, it remains unclear ..." that defines what problem you are  
addressing.  
* A statement of the form:"Here, we ..."  that concisely states your contribution.  
* A series of 3?5 statement about your results.  
* A concluding statement.

# Intro

The first genome sequence of *Arabidopsis thaliana* facilitated rapid advancement of plant biology through molecular genetics. Since this initial genome, massive scaling of sequencing technology has allowed a pioneering group to survey the global genomic variation in natural *Arabidopsis thaliana* populations. This valuable population genetics resource has led to several associations of genetic loci with phenotypic traits and provided insights into how selective pressure has and is influencing the evolution of plant genomes. Outside of genome-wide studies however, this valuable dataset has seen little use. Natural genomic variation also provides a set of feasible functional variation at the gene, gene family and gene network level that may provide insight into the function and evolution of genes, families and networks. Here, we present a web application and R-package through which plant molecular biologists with little-to-no bioinformatics experience can make use of this rich dataset of genetic variation to formulate hypotheses as to the sequence/function/phenotype relationships determined by the gene, family, or network of their interest. We demonstrate the utility of this tool through comprehensive analysis and identification of potential functional variation in the nuclear auxin signaling pathway.

This application allows easy access to 
1. lists of missense polymorphisms to facilitate biochemical assays of variant effects [@starita_variant_2017],
2. family-wise alignments of variants to facilitate de novo functional domain identification [@melamed_combining_2015], 
3. lists of accessions containing missense (or any type of) polymorphisms to facilitate segregation analysis and measurment of the distribution of variant effects on phenotype [@park_distributions_2017], 
4. as well as the study of gene and network evolution [@delker_natural_2010; @kliebenstein_role_2008].

Here, we have utilized this tool to analyze the genetic variation in the nuclear auxin signaling pathway and formulate hypotheses regarding the functional implications of this variation and the evolution of the genes and gene families in this pathway. Our analysis also provided further confirmation of much of the existing knowledge of these genes demonstrating the validity of this approach.

# Methods

## Nucleotide Diversity Calculation
Nei  defined the nucleotide diversity statistic in his original paper as:
“…the average number of nucleotide differences per site between two randomly chosen DNA sequences.” (Nei, 1979)
And provided the equation:

$\displaystyle \pi = \sum_{ij}x_{i}x_{j}\pi_{ij}$

Where Where $x_{i}$ is the frequency of the ith sequence in the population and $\pi_{ij}$ is the number of sites that are different between the ith and jth sequence divided by sequence length.

a more general form, that treats each sequence in the population as unique can be written as:

$\displaystyle \pi = \frac{1}{L*n^2} \sum_{i=1}^n \sum_{j=1}^n \sum_{k=1}^L{\pi_{ijk}}
\qquad
\displaystyle \pi_{ijk} = \left\{\begin{array}{l}
1 \quad \textrm{if} \quad N_{ik} \neq N_{jk} \\
0 \quad \textrm{if} \quad N_{ik} = N_{jk}
\end{array}\right.$

where $N_{ik}$ is the nucleotide (A, T, C or G) at position k on the ith sequence of the population.
$L$ is the length of the sequence (the same for all sequences in the population, 
assumes only SNPs exist, no indels). n is the total number of sequences in the population.

from this form we can re-arrange summations to the form below:

$\displaystyle \pi = \frac{1}{L} \sum_{k=1}^L{\pi_{k}}
\qquad
\pi_{k} = \frac{1}{n^2} \sum_{i=1}^n \sum_{j=1}^n {\pi_{ijk}}$

here $\pi_k$ can be thought of as the site-wise nucleotide diversity at position k, and is equal to 
the nucleotide diversity of a sequence of length 1 at location k. we can calculate $\pi_k$ for each
site, then average those over the sequence length to calculate $\pi$, the nucleotide diversity 
of the sequence.

The function `Nucleotide_diversity` in the r1001genomes package calculates $\pi_k$ for each position in the gene or region that contains a variant. note, $\pi_k$ is equal to 0 at all locations without variants.
This is also what is displayed in the Diversity Plot tab of the web app.

## Detailed $\pi_k$ calculaiton simplification.

The formula for $\pi_k$ above requires comparing every sequence to every other sequence at location k,
however, we know there are only a few variant forms at each individual location.

so we can revert back to using Nei's original formula, modifying it slightly,
replacing $x_i$ with $\frac{n_i}{n}$, $n_i$ being the number of sequences in the population of the form $N_i$ at location k: subscripts i and k are summed only over unique forms [need to clarify this]

$\displaystyle \pi_k = \sum_{ij} \frac{n_i}{n}\frac{n_j}{n} \pi_{ij} = 
\frac{1}{n^2}\sum_{ij} n_i n_j \pi_{ij}$

$\pi_{ij(k)} = \left\{\begin{array}{l}
1 \quad \textrm{if} \quad i \neq j \\
0 \quad \textrm{if} \quad i = j
\end{array}\right.$

we will define $n_{!i} = n - n_i$ as the number of sequences different from i at k.

we can see the summed term will be zero if $i=j$, and $n_i n_j$ if $i\neq j$ so:

$\displaystyle \pi_k = \frac{1}{n^2} \sum_{i} n_i n_{!i}$

next we substitute our definition of $n_{!i}$

$\displaystyle \pi_k = \frac{1}{n^2} \sum_{i} n_i (n - n_{i})$

then distribute and split summation:

$\displaystyle \pi_k = \frac{1}{n^2} (n \sum_{i} n_i - \sum_{i} n_{i}^2)$

finally, summing $\displaystyle \sum_{i}n_i$ is equal to n:

$\displaystyle \pi_k = \frac{1}{n^2} (n^2 - \sum_{i} n_{i}^2)$

this simplified form for $\pi_k$ is used by the app, because the counts of unique 
variants at a single nucleotide location can easily be summarized in R.

# Results  
## *TIR1/AFB* genes

Auxin acts by binding to receptors that in turn target co-repressors for degradation. Auxin receptors (six in the model plant *Arabidopsis thaliana*) evolved through gene duplication and diversification early in the history of vascular plants [@Parry_Complex_2009]. The rate of co-repressor degradation is determined by the identity of both the receptor and co-repressor [@Havens_Synthetic_2012], and this rate sets the pace of lateral root development [@Guseman_Auxin-induced_2015].

Although it is unclear what unique roles each receptor plays in growth and development, a number of studies have pointed out differences in the ways the six different receptors in A. thaliana differ in biochemical function and expression domain[@dharmasiri_plant_2005; @parry_complex_2009; @prigge_arabidopsis_2016]. Although TIR1/AFBs are expressed ubiquitously in A. thaliana tissues, TIR1, AFB2, and AFB3 have been shown to accumulate in the shoot and root meristems and leaf tissues, with slightly different expression patterns for TIR1 [@Parry_Complex_2009]. Additionally, the expression of AFB5 is stronly circadian-regulated [@Covington_Circadian_2007] and AFB3 is more highly expressed in the roots in the presence of nitrate, allowing increased lateral root formation [@Vidal_Nitrate_2010], suggesting more broad environmental regulation of this gene family may exist. 
  Functionally, all members of this family have been shown to bind auxin and Aux/IAA proteins. However, AFB1 has drastically reduced ability to assemble into an SCF complex, due to the substitution E8K in its F-box domain, preventing it from inducing degradation of Aux/IAAs[@Yu_Untethering_2015]. This lack of complexation may allow observed high ubiquitous AFB1 accumulation [@Parry_Complex_2009]. Higher order mutants in the family containing *afb1* mutants suggest that *AFB1* has a moderate positive effect on auxin signaling. Additionally, AFB4 and AFB5 have been shown to preferentially and functionally bind the synthetic auxin picloram, while other family members preferentially bind indole-3-acetic acid [@Prigge_Arabidopsis_2016]. Interestingly, the strength and rate with which TIR1/AFBs are able to bind and mark Aux/IAAs for degradation are variable [@calderonvillalobos_combinatorial_2012; @Havens_Synthetic_2012]. AFB2 induced the degradation of certain Aux/IAA proteins at a faster rate than TIR1, suggesting some functional specificity has arisen since the initial duplication between the *TIR1/AFB1* and *AFB2/AFB3* clades.

```{r TIR1_AFB_div_stats}
gene_info <- geneInfoFromFile(
  fname = system.file("shiny-app", "gene_family_data",
                      "AFB_gene_ids.csv", package = "r1001genomes"))
annotation <- readAnnotationFile(filename = system.file("shiny-app", "gene_family_data",
                      "AFB_annotations.csv", package = "r1001genomes"), gene_info = gene_info)
alignment <- alignCDS(gene_info$transcript_ID)
# make an alignment data frame
aln_df <- makeAlnDF(alignment[[2]])

# make a collapsed alignment for translating positions
collapsed_aln_df <- 
  aln_df %>% dplyr::select(aln_pos, seq_pos, seq_name) %>%
  tidyr::spread(key = seq_name, value = seq_pos)

VCF <- readRDS(system.file("shiny-app", "gene_family_data",
                      "AFBs_VCF.rds", package = "r1001genomes"))
div_stats <- ldply(VCF, diversityStats, geneInfo=gene_info, .id="transcript_ID")
div_stats$tair_symbol %<>% fct_reorder(.x =  div_stats$tair_symbol, 
                                       .fun = function(x) 
                                         {as.integer(str_remove(x, 
                                              pattern = "AFB|TIR"))})
div_stats$tair_symbol %<>% fct_relevel("TIR1")
div_stats %<>% arrange(tair_symbol)
div_stats %<>% select(tair_symbol, everything())

Examining the natural sequence variation across the *AFB* family revealed that *TIR1* and *AFB1* both had very low nonsynonymous diversity, hinting at their likely functional importance and bringing in to question the inconclusive role of *AFB1* in auxin signaling. *AFB3* and *AFB4* had higher nonsynonymous diversity, while their sister genes, *AFB2* and *AFB5* were more conserved. This matches our current understanding of *AFB3* as playing a minor role in the auxin signaling pathway and *AFB4* perhaps undergoing pseudogenization.

Although most known functional regions are highly conserved in *AFB1*, there is a nonsynonymous polymorphism  in the oligomerization domain, only found in the Can-0 accession. Mutations in this domain of *TIR1* frequently have a semidominant effect on root phenotypes [@dezfulian_oligomerization_2016; @wright_insights_2017a]. Characterization of this accession may identify potential *AFB1* phenotypes and may also test the hypothesis that AFB1 functions through oligomerization with other TIR1/AFBs.

```{r AFB variants for PyMOL}
coding_vcf <- ldply(.data = VCF, .fun = subset, 
             !is.na(Transcript_ID) & gt_GT != "0|0")
aln_df <- addSNPsToAlnDF(aln_df, coding_vcf, seq_name = Transcript_ID, 
               seq_pos = Codon_Number)
collapsed_aln_df %<>% left_join(y = subset(aln_df, seq_name == gene_info[gene_info$tair_symbol == "TIR1", "transcript_ID"]))

#' Select variants for importing into PyMOL
#'
#' @param gene_info a gene info data frame
#' @param aln_df an alignment data frame with attached SNPs
#' @param structure_ref string, which tair_symbol or transcript_ID (depending on 
#' \code{name_by_symbols}) corresponds to the protein structure to which 
#' selections should be mapped
#' @param effect_type string of the effect type to select from the alignment 
#' data frame
#' @param name_by_symbols logical, if true (default) tair_symbols will be used 
#' to name the selections, otherwise transcript_IDs will be used
#'
#' @return
#' @export
#'
#' @examples
select_variants <- function(gene_info, aln_df, structure_ref, 
                            effect_type = "missense", name_by_symbols = TRUE){
  collapsed_aln_df <- aln_df %>% dplyr::select(aln_pos, seq_pos, seq_name) %>%
    tidyr::spread(key = seq_name, value = seq_pos)
  if(name_by_symbols)
    column <- "tair_symbol"
    else 
      column <- "transcript_ID"

  selections <- ldply(.data = gene_info[,column], .fun = function(gene){
    collapsed_aln_df %<>% dplyr::left_join(y = subset(aln_df, seq_name ==  
                      gene_info[gene_info[, column] == gene, "transcript_ID"]))
    variants <- collapsed_aln_df[
      grepl(effect_type, collapsed_aln_df$effects), 
      gene_info[gene_info[, column] == structure_ref, "transcript_ID"]]
    c(gene, paste0(na.omit(as.numeric(variants[[1]])), collapse = "+"))
  })
  return(selections)
}

selections <- select_variants(gene_info, aln_df, structure_ref = "TIR1")
selections$V2 <- paste0("chain B and resi ", selections$V2)
write.table(selections, file = "AFB_variant_selections", row.names = FALSE, 
          col.names = FALSE, sep = ",", quote = FALSE)
```

```{python readSelection, eval=FALSE}
#
# -- readSelection.py
#
from pymol import cmd, stored

def readSelection (fileName):
	'''
	DESCRIPTION

	reads a csv file containing a column of selection names and a column of
	selection strings and creates selection objects for each
	'''
	#read in the selection file
	with open(fileName) as f:
		lines = f.readlines()
		for line in lines:
			sel = line.split(",")
			cmd.select(sel[0], sel[1])

	f.close()

cmd.extend('readSelection',readSelection);
```

```{PyMol, AFB_structure, eval=FALSE}
#from within pymol
cmd.load("Dropbox/180705_NIH_renewal/Aim1-figures/2P1Q.pdb")
hide everything

run ~/Desktop/natural-variation-webtool/manuscript/readSelection.py

readSelection ~/Desktop/natural-variation-webtool/manuscript/AFB_variant_selections

# Define selections
# select TIR1variants, (chain B and resi 154+239+546)
select fbox, (chain B and resi 1-50)
select LRR, (chain B and resi 50-576)
select TIR1, (chain B and resi 1-576)
# select AFB2variants, (chain B and resi 38+174+181+184+209+259+401+496)
select tir1_1, (chain B and resi 147)
select TIR1dm, (chain B and resi 170+473)
select insPcontacts, (chain B and resi 74+78+113+114+344+401+403+436+460+484+485+509)
select dimerNULL, (chain B and resi 142+143+151+117+147)
select auxin, resn IAC
select AUXcontacts, (chain B and resi 78+79+82+378+380+403+404+405+438+439+462+463+464+489)
select insP, resn IHP
select IAA7, chain C
select IAA7contacts, (chain B and resi 81+82+84+139+347+349+350+351+352+380+405+406+407+409+464+465+489)
select IAA7_N, chain C and resi 1
select IAA7_C, chain C and resi 13
select ASK1, chain A
select Y, chain C and resi 11
select Ycontacts, chain B and resi 165+166+195

#Define color palette
#set_color name, [ red-float, green-float, blue-float ]
set_color blue1, [0.00000000, 0.20392157, 0.6980392]
set_color blue2, [0.09803922, 0.36078431, 1.0000000]
set_color blue3, [0.00000000, 0.29019608, 1.0000000]
set_color teal,  [0.09803922, 1.00000000, 0.6980392]
set_color LRR,   [1.00000000, 0.71764706, 0.0000000]
set_color Fbox,  [0.69803922, 0.49411765, 0.0000000]
set_color TIR1l, [0.67058824, 0.07843137, 0.8000000]
set_color TIR1f, [0.63529412, 0.34117647, 0.6980392]
set_color AFB2l, [0.37647059, 0.70196078, 1.0000000]
set_color AFB2f, [0.07843137, 0.64313725, 0.8000000]

show cartoon, 2P1Q

color lrr, LRR

color TIR1l, TIR1
show sticks, TIR1
color TIR1f, AFB1
show sticks, AFB1
color AFB2l, AFB2
show sticks, AFB2
color AFB2f, AFB3
show sticks, AFB3

set ray_trace_mode, 3

# used get_view() to return the current view (position) of the molecule 
# copy the code to be able to set the identical viewpoint later
# to return the current view from Pymol
# get_view

set_view (\
     0.044671144,    0.083668359,   -0.995409012,\
     0.774909019,    0.625905454,    0.087390237,\
     0.630409002,   -0.775298953,   -0.036878210,\
    -0.001916260,    0.005584979, -202.498916626,\
    -5.538011074,   -8.507736206,   39.046234131,\
  -219.237945557,  624.123413086,  -20.000000000 )

#to render a high definition image use ray
ray

```{r AFB1 diversity plot}
coding_variants <- getCodingDiv(VCF[[gene_info[gene_info$tair_symbol == "AFB1", "transcript_ID"]]])
p <- plotCodingDiv(uniqueCodingVars = coding_variants)
annotation <- addAlnPosToAnno(anno_df = annotation, aln_df)

# translate positions from TIR1 to AFB1
AFB1_positions <- annotation$positions
AFB1_positions %<>% left_join(y = select(collapsed_aln_df, aln_pos, gene_info[gene_info$tair_symbol == "AFB1", "transcript_ID"]), by = "aln_pos") 
AFB1_positions$position <- AFB1_positions$AT4G03190.1

p <- append_layers(p,list(
        geom_rect(data = AFB1_positions,
                   mapping = aes(xmin = as.integer(position)-1,
                                 xmax = as.integer(position)+2,
                                 fill = annotation),
                  ymin = -Inf, ymax = Inf, inherit.aes = FALSE, alpha = 0.8)),
      position = "bottom")
p
```

The AFB4 and AFB5 receptors have an N-terminal extension prior to the F-box domains. This extension had very high nonsynonymous diversity(Fig. 4), suggesting that this extension does not play an important functional role in these proteins. Additionally, two frameshift variants and one stop-gained variant were observed in *AFB4* supporting its pseudogenization.

Figure 4: Alignment of A. thaliana TIR1/AFB F-box domains showing variants. Cyan marks the F-box binding domain. Yellow, green, and magenta, mark missense and synonymous, synonymous only, and missense only variant locations respectively. Yellow squared E’s show Cul1-interacting residues, differing in AFB1.
```{r TIR1/AFB F-box Alignment}
aln_df <- makeAlnDF(alignment[[2]])
aln_df <- left_join(aln_df, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
aln_df$tair_symbol <- fct_rev(fct_relevel(aln_df$tair_symbol, "TIR1"))
aln_df <- chunkAlnDF(aln_df = aln_df, chunk_width = 55)
VCF <- ldply(.data = VCF, .fun = subset, gt_GT != "0|0")
VCF <- getCodingDiv(VCF)
aln_df <- addSNPsToAlnDF(aln_df, VCF)
F_box <- subset(aln_df, aln_pos <= 96)
F_box <- chunkAlnDF(aln_df = F_box, chunk_width = 80)
p <-ggplot(F_box, aes(x = aln_pos, y = tair_symbol,
                             group = seq_pos, text = variants))
p <- p +
  geom_tile(data = na.omit(F_box), mapping = aes(fill = effects),
            width = 1, height = 1, alpha = 0.8) +
  geom_text(aes(label=letter), alpha= 1, family = "Courier") +
  scale_fill_brewer(type = "qual", palette = 1, direction = -1) +
  scale_x_continuous(breaks=seq(1,max(F_box$aln_pos), by = 10)) +
  scale_y_discrete() +
  # expand increases distance from axis
  labs(x = "", y = "", fill = "variant") +
  theme_logo(base_family = "Helvetica") +
  theme(panel.grid = element_blank(), panel.grid.minor = element_blank()) +
  facet_wrap(facets = ~chunk, ncol = 1, scales = "free") +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank(),
        legend.box = "vertical")
p
```

## *Aux/IAA* genes

@@

## *TPL/TPR* genes

The Auxin signaling pathway utilizes the TOPLESS (TPL) and TOPLESS-related (TPR) family of Gro/TLE/TUP1 type co-repressor proteins to maintain auxin responsive genes in a transcriptionally-repressed state in the absence of auxin [@szemenyei_topless_2008]. In *Arabidopsis thaliana* the five member *TPL/TPR* family includes *TPL* and *TPR1-4*. The resulting proteins are comprised of three structural domains: an N-terminal TPL domain and two WD-40 domains [@long_topless_2006]. TPL/TPR proteins are recruited to the AUX/IAA proteins through interaction with the conserved Ethylene-responsive element binding factor-associated amphiphilic repression (EAR) domain [@szemenyei_topless_2008]. Canonical EAR domains have the amino acid sequence LxLxL, as found in most AUX/IAAs (figure xA) [@overvoorde_functional_2005]. TPL/TPR co-repressors bind EAR domains via their C-terminal to LisH (CTLH) domains found near their N-termini (citations of pre-structure founding papers/reviews). Recent structural analyses of the TPL N-terminal domain have highlighted the precise interaction interface between TPL and AUX/IAA EAR domains, as well as the TPL-TPL dimerization and tetramerization motifs [@martin-arevalillo_structure_2017; @ke_structural_2015a]. The residues required for higher-order multimers of TPL tetramers have also been identified [@ma_d53_2017]. Additional interactions with transcriptional regulation and chromatin modifying machinery are likely mediated by two tandem beta propeller domains of TPL/TPRs. These domain have not yet been crystallized, however numerous high resolution crystal structures of beta propeller domains allow for structural prediction of the TPL C-terminal beta propeller domains and the locations of likely interaction faces (Figure, X).

The TOPLESS co-repressor family generally exhibits a high level of sequence conservation at the amino acid sequence level across resequenced *Arabidopsis thaliana* accessions, with all ${\pi_N/\pi_S}$ values below 1 (Table X). The closely related *TPL* and *TPR1* have the highest ${\pi_N/\pi_S}$ values (TPL-0.425, TPR1-0.739), suggesting that these these two related genes tolerate a higher degree of sequence and potentially functional diversity compared to *TPR2/3/4*. The N-terminal TPL domain of the TPL/TPR family is particularly conserved (Figure X). All nonsynonymous polymorphisms observed in this region are either in the coils between helices or are highly conservative mutations within helices (i.e. Valine to Isoleucine), which would be predicted to exhibit little effect on folding and function.

```{r TPL SNP stats, fig.height=11}
gene_info <- geneInfoFromFile(
  fname = system.file("shiny-app", "gene_family_data",
                      "TPL_gene_ids.csv", package = "r1001genomes"))
#annotation <- readAnnotationFile(filename = system.file("shiny-app", "gene_family_data",
#                      "IAA_annotations.csv", package = "r1001genomes"), gene_info = gene_info)

VCF <- readRDS(system.file("shiny-app", "gene_family_data",
                      "TPL_VCF.rds", package = "r1001genomes"))
div_stats <- ldply(VCF, diversityStats, geneInfo=gene_info, .id="transcript_ID")
div_stats$tair_symbol %<>% fct_reorder(.x = , div_stats$tair_symbol, 
                                       .fun = function(x) 
                                         {as.integer(str_remove(x, 
                                              pattern = "TPR"))})
div_stats$tair_symbol %<>% fct_relevel("TPL")
div_stats %<>% arrange(tair_symbol)
div_stats %<>% select(tair_symbol, everything())

The high degree of conservation in the entire N-terminal domain underscores its importance in TPL/TPR function. For example, the initial *tpl-1* mutation (N176H) in the ninth helix is a dominant gain-of-function allele [@long_topless_2006], which is capable of binding wild-type TPL protein and inducing protein aggregation [@ma_d53_2017]. It is therefore understandable that this helix had very low diversity as nonsynonymous variants in this domain could act in a dominant negative fashion. Addtionally, the crystal structure of TPL has revealed that the The LisH domain and CRA domain coordinate TPL dimerization and tetramerization respectively (26601214, 28698367), the CTLH domain is critical for interaction with DNA-binding transcription factors (26601214), and helix nine within the CRA domain may be critical for the formation of higher-order TPL complexes (multimers of TPL tetramers, 28630893).

Compared to the TPL domain, both the proline rich and WD-40 domains exhibited a higher frequency of missense variants. We utilized Phyre2.0 modeling software to predict the structure of the TPL/TPR WD-40 repeats in order to map non-synonymous mutations to protein structure (Figure X). Here we can see both faces of each predicted beta-propeller structure of the WD-40 domains, which are connected by a short peptide linker. This model reveals amino acid conservation in the core and central residues of the WD-40 domains. ~~These observations could be used as the basis for forward mutagenesis strategies to test whether this region of TPL functions similarly to its homolog TLE, which binds its target genes EH1 and the C-terminal WRPW/Y motif (Hairy/Hes/Runx) in the center of the WD-40 domain (pmid 16762837).~~ **Can we say something more general here? There is not enough background to support this. What is TLE from? What is EH1? Is TLE really binding to EH1 DNA?**

```{r TPL_TPR alignment}

Despite the presence of five *TPL/TPR* genes in the genome, which are considered to be redundant (pmid 16763149), very few observed SNPs are predicted to result in loss of function for any *TPL/TPR* gene. Only one frameshift variant was observed, *TPR4-665delG*, in a single accession (Aiell-4 from Italy). This deletion occurs in codon 222, just after the TPL N-terminal domain, and generates an early stop eleven codons downstream from the deletion. It is unclear whether this N-terminal fragment would be expressed at any appreciable level, as it is possible that the generation of an early stop codon would trigger nonsense-mediated decay. However, if expressed, it would be fascinating to examine whether this naturally occurring N-terminal truncation retains activity as a co-repressor. Given the observation that the TPL N-terminal domain is sufficient to bind histones (28630893), and repress transcription via synthetic assays in yeast (24979769), it is possible that this represents a semi-functional TPL variant. There has been no evidence for prioritization of individual *TPL/TPR* family members in the auxin response pathway, but experiments modulating the gene copy number of *TPR1* revealed the importance of maintaining the proper dosage of *TPL/TPR* in order to accurately trigger the salicylic acid innate immune response (20647385). These results, taken together with the absence of *TPL/TPR* loss-of-function alleles in the *Arabidopsis thaliana* accessions suggest that perhaps the entire family is being conserved at the nucleotide level in the population to meet a gene dosage requirement, with little opportunity for neo- or sub-functionalization.

## ARF genes

Auxin response is ultimately mediated by the auxin responsive transcription factors (ARFs). There are 23 ARFs in *Arabidopsis thaliana* that are divided into three phylogenetic classes. Class A ARFs (ARF5, ARF6, ARF7, ARF8 and ARF19) activate transcription. These ARFs have a glutamine-rich region in the middle of the protein that may mediate activation (Guilefoyle, 2007). It has recently been shown that the middle region of ARF5 interacts with the SWI/SNF chromatin remodeling ATPases BRAMA and SPLAYED, possibly to reduce nucleosome occupancy and allow for the recruitment of transcription machinery (Wu, 2016). Additionally, ARF7 interacts with Mediator subunits, directly tethering transcriptional activation machinery to its binding sites in the chromosome (Ito, 2016). Class B and C ARFs are historically categorized as repressor ARFs, though the mechanism through which they confer repression has not been identified. Their middle regions tend to be proline- and serine-rich (citation, Ulmasov?).

Canonical ARFs are comprised of three major domains. Recent crystallization of these domains have informed structure-function analysis of the ARFs (Boer, 2014; Nanao, 2014; Korasick, 2014). These domains are conserved throughout land plants (Dolf's new paper). ARFs share an N-terminal B3 DNA binding domain. Flanking this DNA-binding domain is a dimerization domain, which folds up into a single “taco-shaped” domain to allow for dimerization between ARFs. There is an auxiliary domain that immediately follows and interacts with the dimerization domain. The middle region is the most variable between ARFs, as mentioned above, but is characterized by repetitive units of glutamine (class A), serine, or proline residues (classes B and C).

The C-terminal domain of the ARF is a protein-protein interaction domain mediating interactions among ARFs, between ARFs and other transcription factors, and between ARFs and the Aux/IAA repressors. This interaction domain was recently characterized as a Phox and Bem1 (PB1) domain, which is comprised of a positive and negative face with conserved basic and acidic residues, respectively (Nanao, 2014; Korasick, 2014). The dipolar nature of the PB1 domain may mediate multimerization by the pairwise interaction of these faces on different proteins as the ARF7 PB1 domain was crystallized as a multimer [@korasick_]. However, it is unclear whether ARF multimerization occurs or plays a developmental significant role in vivo. Interfering with ARF dimerization in either the DNA-binding proximal dimerization domain or the PB1 domain decreases the ability of class A ARFs to activate transcription in a heterologous yeast system (Pierre-Jerome, 2016).

While domain architecture is broadly conserved among the ARFs, there are exceptional cases. Three ARFs do not contain a PB1 domain at all, ARF3, ARF13, and ARF17, and several more have lost the conserved acidic or basic residues in the PB1 domain, suggesting they may be reduced to a single interaction domain. Several ARFs additionally have an expanded conserved region within the DNA-binding domain, of unknown function. The majority of domain variation among ARFs occurs in the large B-class subfamily. The liverwort *Marchantia polymorpha* has a single representative ARF of each class (Flores-Sandoval, 2015). The expansion of these classes in flowering plants is the result of both whole genome and tandem duplication events (Remington, 2004). The growth of the ARF family may have allowed for the expansion of the quantity and complexity of loci regulated by the ARFs and subsequent expansion in their regulation of developmental processes.

Class A ARFs are the most well-studied ARF subfamily---the five family members all act as transcriptional activators and have well-characterized, distinct developmental targets. Overall the diversity of class A ARFs was generally low, especially compared to the class B and C ARFs, suggesting that class A ARFs are central to auxin signal transduction and plant development. Analysis of class A ARF nonsynonymous diversity suggests that the majority of these ARFs are highly functionally conserved, with ${\pi_N/\pi_S}$ values much lower than 1 with the exception of ARF19, with ${\pi_N/\pi_S}$ value of 1.8. Comparing diversity within sister pairs, there is a similar trade-off as seen in most *IAA* sister pairs, with one sister being highly conserved and the other more divergent. ARF19 and ARF8 are the more divergent class A ARFs, with ${\pi_N/\pi_S}$ values at least three time those of their sisters, ARF7 and ARF6 respectively. This may suggest that ARF6 and ARF7 serve more essential purposes in plant development.

```{r ARF diversity stats}
gene_info <- geneInfoFromFile(
  fname = system.file("shiny-app", "gene_family_data",
                      "ARF_gene_ids.csv", package = "r1001genomes"))
#annotation <- readAnnotationFile(filename = system.file("shiny-app", 
  #"gene_family_data", "ARF_annotations.csv", 
  #package = "r1001genomes"), gene_info = gene_info)

VCF <- readRDS(system.file("shiny-app", "gene_family_data",
                      "ARF_VCF.rds", package = "r1001genomes"))
div_stats <- ldply(VCF, diversityStats, geneInfo=gene_info, .id="transcript_ID")
div_stats$tair_symbol %<>% fct_reorder(.x =  div_stats$tair_symbol, 
                                       .fun = function(x) 
                                         {as.integer(str_remove(x, 
                                              pattern = "ARF"))})
div_stats %<>% arrange(tair_symbol)
div_stats %<>% select(tair_symbol, everything())

```{r ARF alignment}
alignment <- alignCDS(gene_info$transcript_ID)
# make an alignment data frame
aln_df <- makeAlnDF(alignment[[2]])

```{r Mask_ARF_alignment, eval=FALSE}
aliscore(as.AAbin(alignment[[2]]), path = "/Volumes/SSD/Applications/ALISCORE_v2.0/Aliscore_v.2.0/")
id <- scan(file = "/Volumes/SSD/Applications/ALISCORE_v2.0/Aliscore_v.2.0/input.fas_List_random.txt", sep = " ", quiet = TRUE)
AAAlign_no_gaps <- as.matrix(alignment[[2]])[, id]
write.nexus.data(AAAlign_no_gaps, file = "ARFs.nex", format = "protein")
```
```{bash ARF_phylogeny_construction, eval=FALSE}
/Applications/MrBayes/mb

execute ARFs.nex
# set ARF17 as the outgroup
outgroup AT1G77850.1

```{r Load ARF phylogeny and data}
tree <- treeio::read.beast(file = "ARFs.nex.con.tre")

ARF_data <- read.csv(system.file("shiny-app", "gene_family_data",
                      "ARF_gene_ids.csv", package = "r1001genomes"))
ARF_data %<>% left_join(y = div_stats$x$data, by = c("name" = "tair_symbol"))
#tip_labels need to be in the first column
ARF_data %<>% select(name, everything())
ARF_data$tair_locus %<>% as.character()

ARF_data <- ARF_data[match(gsub(x = tree@phylo$tip.label, pattern = "\\.1",replacement = ""), ARF_data$tair_locus), ]

tree@phylo$tip.label <- ARF_data$name

```{r Visualize ARF tree, fig.height=5, fig.width=5}
p <- ggtree(tree, layout = "rectangular", color = "grey") %<+% ARF_data + theme_tree() + scale_x_continuous(expand = c(0, 1.2)) + geom_tiplab(hjust = -0.25) + 
  geom_tippoint(mapping = aes(size = 4*Pi_NS_Ratio, fill = Pi_NS_Ratio), 
                shape = 21, color = "white", show.legend = TRUE) + 
  scale_fill_viridis(values = rescale(c(min(ARF_data$Pi_NS_Ratio), 1,
                                        max(ARF_data$Pi_NS_Ratio)), 
                                      to = c(0,1)), guide = "colourbar") + scale_size(guide = "none") + 
  theme(legend.position = c(.1,.7)) + 
  geom_nodelab(mapping = aes(label = round(prob,digits = 2)), nudge_x = -.3, nudge_y = .5, size = 3) +
  geom_tiplab(mapping = aes(color = class), hjust = -0.25) + 
  scale_color_viridis(discrete = TRUE)
p
```

### Class A ARFs 
For all class A ARFs, the middle region of the protein was the predominant high diversity region. The DNA-binding domain had very few, low-diversity missense mutations, as did the PB1 domain. Considering the necessity of their conserved functions, the low level of variation in these key DNA and protein-protein interaction domains is expected.

>>>>

# *Aux/IAA* genes {#IAAs}

@@
